Exploratory Data Analysis of Known Exoplanets by Luke Tibbott
========================================================




This data set was retrieved from the NASA Exoplanet Archive. An exoplanet is an
extrasolar planet -- that is, a planet that exists outside of earth's solar 
system. The original data set has been pared down to 17 columns from 140+
columns. 

When cleaning this dataset I was confronted with the issue of what to do with
the large amount of null values in certain columns. Removing all the rows with 
nulls results in a tiny dataset, and imputing the values is very dangerous due
to the possibility of assigning incorrect values. Most null values are due to 
how a particular data point was captured -- different exoplanet discovery 
missions are simply incapable of capturing certain characteristics of an 
exoplanet. Nonetheless, any good data we can add to our dataset is valuable
to our analysis, so I decided to keep the null values.




```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report


library(tidyverse)


```

```{r echo=FALSE, Data_cleaning}
# Read in data to a dataframe object
df <- read.csv('planets.csv', header=TRUE)

# Grab columns 2 through 20
pa <- df[, 2:20]

pa$pl_name <- df$pl_name
pa$st_optmag <- df$st_optmag
pa$st_dist <- df$st_dist
pa$st_spster <- df$st_spstr
pa$st_age <- df$st_age
pa$st_mass <- df$st_mass
pa$st_dist <- df$st_dist
pa$st_teff <- df$st_teff
pa$pl_bmasse <- df$pl_bmasse

# Drop columns we don't care about
ep <- pa[ , -which(names(pa) %in% c('pl_orbeccen', 'pl_orbincl', 
                                    'pl_ttvflag', 'pl_kepflag',
                                    'pl_k2flag', 'pl_nnotes',
                                    'ra', 'dec_str', 'pl_bmassprov',
                                    'ra_str'))]


```



```{r echo=FALSE, Data_transformation}

# Reorder the discovery method factor according to the count of planets
# discovered using each method, descending.
ep$pl_discmethod <- factor(ep$pl_discmethod, levels = 
                             c('Transit','Radial Velocity', 'Microlensing',
                               'Imaging', 'Transit Timing Variations',
                               'Eclipse Timing Variations', 
                               'Orbital Brightness Modulation',
                               'Pulsar Timing', 'Pulsation Timing Variations',
                               'Astrometry'))

# Change the spectral type column to only be the first letter of the string.
# This will give us a broad definition of the star's spectral type rather than
# a very specific one
ep$st_spster <- substring(ep$st_spster, 1, 1)

# Factor spectral type column to ensure ggplot plays nice with it
ep$st_spster <- factor(ep$st_spster, 
                       levels = c('A', 'B', 'F', 'G', 'K', 'L', 'M', 'S', 
                                  'T', 'W'))

# Here we essentially exclude all planets that are not the "final" planet in
# their system, and use this dataframe to keep track of the number of planets
# in each system
num_planets = subset(ep, (ep$pl_letter == 'b' & ep$pl_pnum == 1) |
                       (ep$pl_letter == 'c' & ep$pl_pnum == 2) |
                       (ep$pl_letter == 'd' & ep$pl_pnum == 3) |
                       (ep$pl_letter == 'e' & ep$pl_pnum == 4) |
                       (ep$pl_letter == 'f' & ep$pl_pnum == 5) |
                       (ep$pl_letter == 'g' & ep$pl_pnum == 6) |
                       (ep$pl_letter == 'h' & ep$pl_pnum == 7) | 
                       (ep$pl_letter == 'i' & ep$pl_pnum == 8))

# These dataframes consist of observations of each of the four discovery methods
trans <- ep %>% subset(pl_discmethod == 'Transit')
rv <- ep %>% subset(pl_discmethod == 'Radial Velocity')
ml <- ep %>% subset(pl_discmethod == 'Microlensing')
im <- ep %>% subset(pl_discmethod == 'Imaging')

# Subset the dataframe to ignore less common discovery methods which account for
# a very small portion of the dataset
ep <- subset(ep, pl_discmethod == 'Transit' | pl_discmethod == 'Radial Velocity' | 
           pl_discmethod == 'Microlensing' | pl_discmethod == 'Imaging')

```



```{r echo=FALSE, Functions}
# Produces a histogram plot using ggplot2.
# Arguments: df, a dataframe object; x, a variable in df;
# binwidth, width of the bars in the histogram

hist_plot <- function(df, x, binwidth) {
  ggplot(df, aes(x = x)) + geom_histogram(binwidth=binwidth)
}

# Produces a scatter plot using ggplot2.
# Arguments: df, a dataframe object; x, a variable in df, y a variable in df

scatter <- function(df, x, y) {
  ggplot(df, aes(x = x, y = y)) + geom_point()
}

```

# Univariate Plots Section

```{r echo=FALSE, Summarize_dataframe}
summary(ep)
```

```{r echo=FALSE, structure_of_dataframe}
str(ep)
```

This dataset has 17 variables and about 3700 observations.

```{r echo=FALSE, Univariate_Plots}

# Rotate the x-axis text so each bar's label is readable
ggplot(ep, aes(x = pl_discmethod)) + geom_bar() + 
  theme(axis.text.x =  element_text(angle = 70, size = 8, hjust = 1))

```

This bar chart shows us that transit and radial velocity methods of discovery
are far and way the most popular methods for discovering exoplanets.
Each of the other methods account for a very small proportion of all discovered
exoplanets.

A couple notes and questions regarding the 'discovery method' variable analyzed
in this project:

* The transit method relies on an exoplanet passing between Earth and a host
 star. The exoplanet blocks some of the star's electromagnetic (EM) energy, thus 
 confirming its existence. Perhaps a planet with a larger radius would block 
 more EM energy and make it easier to discover?
 
* The radial velocity method looks at small changes in a star's motion (eg
whether the star is moving towards earth or away from Earth) to infer the
existence of a planet. I wonder if more massive planets affect the motion of 
stars more, making it easier to discover more massive planets with this method?

* Microlensing relies on a magnification effect from a massive and very distant
 object to allow the observation of objects that don't emit much electromagnetic
 energy, like a planet. What planetary attribute would lend itself to more 
 planets being discovered this way?
 
* Imaging relies on us being able to directly see a planet. Perhaps planets 
 are very close are discovered using imaging more frequently.
 
* The six other methods of discovering planets will not be explored in this
 analysis.

```{r echo=FALSE, warning=FALSE, message=FALSE}

hist_plot(ep, ep$pl_bmassj, .5) + xlab('Jupiter masses')

```

This plot shows us that a large number of planets are very close to
0 Jupiter masses (although none of them are exactly 0). This would lead me to
believe these planets are not gas giants with huge mass and huge radius, but
perhaps small, rocky planets. I would guess the planets with higher mass are gas
giants. We can take a look at planet radius later on to see if these planets'
radii match up with the large numbers we'd expect from gas giants.


```{r echo=FALSE, warning=FALSE, message=FALSE}

hist_plot(ep, ep$pl_bmassj, .25) + coord_cartesian(xlim = c(0, 10)) + 
  scale_y_log10() + xlab('Jupiter Masses')

```

The log-10 adjusted graph with a smaller x-axis gives us a better idea of how
planetary mass is distributed near 0 Jupiter masses. Again, we see a heavily
right skewed distribution. 

```{r echo=FALSE, warning=FALSE, message=FALSE}

hist_plot(ep, ep$pl_radj, .1) + xlab('Jupiter radii')

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

hist_plot(ep, ep$pl_radj, .1) + xlab('Jupiter radii') + 
  coord_cartesian(xlim = c(0, 2.5))

```


Most of the planets are less than .25 Jupiter radii, but there is a small
peak a little over 1 Jupiter radius. I expected the planets' radii to be be much
higher after seeing the mass plot. This makes me think that many of the planets
have a high density.

```{r echo=FALSE, warning=FALSE, message=FALSE}

hist_plot(ep, ep$pl_dens, 1) + xlab('Density, g / cm^3')

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

hist_plot(ep, ep$pl_dens, .5) + xlab('Density, g / cm^3') + 
coord_cartesian(xlim = c(0, 20))

```


As suspected, we do see a bit more variance in density, but the relationship
between mass, radius, and density needs to be investigated more. Unfortunately
we have significantly less data on density than on mass and radius.

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(ep, aes(x = pl_bmassj)) + geom_histogram() +
  facet_wrap(~ep$pl_discmethod) +
  coord_cartesian(xlim = c(0, 30)) + scale_y_log10() +
  ggtitle('Planetary mass distribution by discovery method') + 
  theme(plot.title = element_text(hjust = 0.5))

```



* Note that all y axes are scaled by log 10 to make the histograms for the 
microlensing and imaging discovery methods visible.

* The mass of planets discovered via the transit method is heavily concentrated
below 2.5 Jupiter masses, with relatively low variance.

* Planets discovered via the radial velocity method have a distribution very
similar to those discovered via transit.

* We have very few data points for planets discovered via microlensing, but 
this distribution has much higher variance. Most planets discovered via
microlensing have relatively low mass, but a significant proportion of planets
discovered via microlensing have more than 2.5 Jupiter masses.

* We have even less data for planets discovered via imaging, but this 
distribution is the only one we've seen that isn't clustered near 0 Jupiter 
masses. I would guess that it's easier to discover more massive planets via 
imaging because they are bigger and easier to see. Let's take a look at
the radius of planets separated by discovery method.

```{r echo=FALSE, warning=FALSE, message=FALSE}

hist_plot(ep, ep$pl_radj, .1) +
  labs(x = 'Jupiter radii', title = 'Planetary radius by discovery method') + 
  scale_y_log10() + 
  facet_wrap(~ep$pl_discmethod) + theme(plot.title = element_text(hjust = 0.5))

```

All discovery methods have distributions almost entirely less than 4 Jupiter
radii. This is surprising, and seems to suggest that my theory that it's easier
to discovery planets with larger radii is wrong. Again, all y axes are scaled by
log 10.


```{r echo=FALSE, num_planets}

ggplot(num_planets, aes(x = factor(pl_pnum))) + geom_bar()


```

Most planetary systems we've discovered have 
only one planet. I'm interested to see if certain discovery methods lend
themselves to discovering systems with more or less planets in them.

```{r echo=FALSE, warning=FALSE, message=FALSE}

hist_plot(ep, ep$st_age, .1) + xlab('Star age (billions of years)')

summary(ep$st_age)

```



The distribution of parent stars is certainly unimodal although perhaps not
quite normal, with a clear peak a bit older than 4 billion years old, and a
higher concentration of stars between 0 and 4.5 billion years old than stars 
between 4.5 and 12 billion years old. This could indicate a number of things,
including our current detection methods being biased towards finding planets
orbiting younger stars, or perhaps younger stars are more likely to host
planets.

A five-number summary of star age shows us the median and mean are both just
over 4 billion years, the interquartile range is 3 to 4.725 billion years. The 
minimum is .001 billion years and the max is 12.5 billion years. This means
we have discovered planets orbiting both extremely young and extremely old
stars, but most are orbiting stars whose age lies somewhere in between. 

```{r echo=FALSE, warning=FALSE, message=FALSE}

hist_plot(ep, ep$st_mass, 1) + xlab('Solar masses')

```


```{r echo=FALSE, warning=FALSE, message=FALSE}

hist_plot(ep, ep$st_mass, .1) + coord_cartesian(xlim = c(0, 5)) + 
  xlab('Solar masses')

```

The stars's masses in our dataset form a normal distribution around 1 solar mass.

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(subset(ep, ep$st_spster != ''), aes(x = st_spster)) +
  geom_bar()

```

Most stars with exoplanets orbiting them can be lumped in to one of four
stellar classifications: F, G, K, or M.

Interestingly, the most common stellar
classification is G, which is the same classification as our own sun.

Class M stars are the most common stars in the universe, so it is not
surprising that many stars in our dataset are class M -- although it is
worth noting that despite class M stars prevalence in the universe, they are not
the most common stars that host exoplanets (that we've discovered). 

Class K stars are an intermediary class between G and M stars.

Class F stars are similar in mass to the Sun, but emit more UV radiation. 

```{r echo=FALSE, warning=FALSE, message=FALSE}

hist_plot(ep, ep$st_teff, 100) + xlab('Effective temperature (K)')

```


Effective temperature of stars in our dataset form an approximately normal
distribution around 6000 K, with a slight skew to the left.


# Univariate Analysis

### What is the structure of your dataset?

There are 3696 observations and 16 variables in this dataset. Not all variables
have 3696 observations -- there are many nulls. Regardless, there are 3696 
planets represented here. 

Over 75% of them were discovered via the transit method. Most planets are part
of a one planet system. Most planets have between 1 and 2 Jupiter masses, and 
between 0 and .5 Jupiter radii. 

Most stars these planets orbit are less than 1000 light years away. Most host
stars weigh the same or less than the sun and are less than 5 billion years old. 
In fact, most of the host stars are between 4 and 5 billion years old.

### What is/are the main feature(s) of interest in your dataset?
Discovery method is perhaps the focal point of this dataset. The transit 
discovery method seems to have been the most fruitful exoplanet discovery
method, which means it's important to do some analysis of this discovery
method. Are there certain planets that this method of discovery is biased
toward discovering? If so, what are the features of those planets? 

It's also possible that this method of discovery is *not* biased, but is simply
discovering a sample distribution of planets representative of *all* planets
that exist in the universe, although testing this notion might be beyond the
scope of our analysis.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The various attributes of the stars in the dataset will help investigate the 
discovery methods. It's possible that certain star types are predisposed to 
certain discovery methods.

Planetary attributes such as mass, density, and radius will give us a good idea
of the types of planets each discovery method is more inclined to discover. 
These attributes can also be compared to attributes of stars to see any 
relationship that might exist there.

### Did you create any new variables from existing variables in the dataset?

I created a new dataframe which tracks the number of planets in a system. A 
histogram of the column pl_pnum is not representative of the number of planets
in a system because every planet in a system of say 5 planets would have 5 
listed as the number of planets in a system. This would mean that a system of
5 planets would add 25 to the histogram plot, rather than just 5.

I attempted to impute some of the null density values using the volume and 
radius columns, but it turned out that all rows with null density values also
have either a null volume or null radius value as well, making density 
impossible to compute.

### Of the features you investigated, were there any unusual distributions? 

I'm surprised by the number of 'big' planets in general! Although histograms 
show a lot of data points near zero, there are still a huge number of data
points above 1 Jupiter mass. Jupiter is the *largest* planet in our solar
system, which makes it so surprising to see such a large number of planets above
1 Jupiter mass. This makes me think that large planets are generally easier to
discover than smaller planets.

### Did you perform any operations on the data to tidy, adjust, or change the \
form of the data? If so, why did you do this?

I grabbed the first character from the spectral type column. This is because
many stars in the dataset were described too precisely for our needs, ie a
star might have been listed as type 'K-IIIbv' which is a very narrow 
sub-classification of a type K star. All we care about is the first letter 
in this column.

I dropped over 100 columns from the dataframe, and created a new dataframe that 
lists the number of planets in a system without repeating this data as it is 
in our uncleaned dataframe. I also dropped all data points that weren't 
collected via the transit, radial velocity, imaging, or microlensing discovery
methods because there was very little data for these discovery methods.


# Bivariate Plots Section

```{r echo=FALSE, warning=FALSE, message=FALSE, Bivariate_Plots}

scatter(ep, ep$pl_dens, ep$pl_bmassj) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 10))

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

scatter(ep, ep$pl_dens, ep$pl_bmassj) +
  coord_cartesian(xlim = c(0, 15), ylim = c(0, 5)) +
  labs(x = 'Density (g/cm^3)', y = 'Jupiter masses')

```


The line of points hugging 0 mass with density between 0 and 15 g/cm^3 is 
most likely representing terrestrial planets. These planets would have high
density due to their rocky composition, but low mass due to size limits (a 
terrestrial planet has a theoretical size limit of about 2 *earth* radii
according to [this study](https://arxiv.org/pdf/1311.0329v1.pdf)). The second
plot clearly shows two different groups of planets.

```{r echo=FALSE, warning=FALSE, message=FALSE}

scatter(ep, ep$pl_radj, ep$pl_bmassj) + 
  coord_cartesian(xlim = c(0, 3), ylim = c(0, 30)) +
  labs(x = 'Jupiter radii', y = 'Jupiter masses')

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(ep, aes(x = pl_radj, y = pl_bmassj)) + geom_point(alpha = 0.2) +
coord_cartesian(xlim = c(0, 3), ylim = c(0, 10)) +
  labs(x = 'Jupiter radii', y = 'Jupiter masses')


```

Mass and radius have less of a positive correlation than I would expect. By 
changing the alpha of points in this plot we can see the large grouping of
what I believe are terrestrial planets near 0 Jupiter radii and 0 Jupiter 
masses. The gaseous planets appear to have a large grouping near 1 
Jupiter radius and 1 Jupiter mass, yet a significant number of planets have
huge mass that doesn't follow the trend of other gas planets. This makes me 
think there are a several distinct groups of planets in our data set. Could
this be due to the bias of different discovery methods to discover different 
types of planets?

```{r echo=FALSE, warning=FALSE, message=FALSE}

scatter(ep, ep$pl_orbsmax, ep$pl_bmassj) + 
  coord_cartesian(xlim = c(0, 8), ylim = c(0, 10)) +
  labs(x = 'Orbital distance (Astronomical Units)', y = 'Jupiter masses')

```


There's no obvious correlation between a planet's max orbit and a planet's
mass, but there is an interesting vertical line very close to 0 astronomical
units. These are likely gas planets orbiting very close to their star (known as
hot Jupiters).

```{r echo=FALSE, warning=FALSE, message=FALSE}

scatter(ep, ep$st_dist, ep$pl_bmassj) +
labs(x = 'Star distance(light years)' , y = 'Jupiter masses') +
  coord_cartesian(ylim = c(0, 30))

```

There appears to be a weak negative correlation between star distance and 
planet mass. This is somewhat counterintuitive because I would expect higher
mass planets to be easier to discover than low mass planets particularly
when the host star is far away. Perhaps this is explained by limitations in 
planetary discovery methods at different distances.



```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(subset(ep, ep$st_spster == 'F' | ep$st_spster == 'G' |
                ep$st_spster == 'K' | ep$st_spster == 'M'), 
              aes(x = st_spster, y = st_dist)) + geom_boxplot() + 
  coord_cartesian(ylim = c(0, 1000))

```

All stars with planets around them have an interquartile range near 0 with
respect to distance from Earth. It seems that more planets have been discovered
orbiting stars near Earth than far from Earth across the four most common
star classifications in our dataset. All stellar classifications have a third 
quartile at or less than 250 light years.

```{r echo=FALSE, warning=FALSE, message=FALSE}

scatter(ep, ep$st_mass, ep$pl_bmassj) + 
  coord_cartesian(xlim = c(0, 15), ylim = c(0, 30)) + 
  labs(x = 'Stellar mass in solar masses',
       y = 'Planetary mass in Jupiter masses') 

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

scatter(ep, ep$st_mass, ep$pl_bmassj) + 
  coord_cartesian(xlim = c(0, 2), ylim = c(0, 5)) +
  labs(x = 'Stellar mass in solar masses',
       y = 'Planetary mass in Jupiter masses')

```

Most stars have mass around 1 solar mass or less. Planetary mass does not
appear to be correlated to star mass.


```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(num_planets, aes(x = factor(pl_pnum), y = st_mass)) + geom_boxplot() +
  coord_cartesian(ylim = c(0, 2.5))

```

There doesn't appear to be any significant difference in the mass of a star
when compared to the number of planets orbiting a star.

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Subset the dataframe to ignore stellar classifications with very little or
# null data

ggplot(subset(ep, ep$st_spster != 'W' & ep$st_spster != 's' &
                ep$st_spster != 'L' & ep$st_spster != 'B' & ep$st_spster != ''), 
       aes(x=factor(st_spster), y = st_teff)) + geom_boxplot() +
  coord_cartesian(ylim = c(0, 10000))

```

Stars classified later in the alphabet tend to haved lower effective 
temperatures among stars that have exoplanets orbiting them. This likely
has something to do with how stars are classified.

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Set the alpha to 0.05 to show where the data is overplotted

ggplot(ep, aes(x = st_mass, y = st_teff)) + geom_point(alpha = 0.05) + 
  coord_cartesian(ylim = c(0, 12000), xlim = c(0, 2))

```

There's correlation between star mass and star effective temperature. I 
hesitate to call it strong, but we definitely see tightly packed observations
trending upward linearly. This plot is reminiscent of a sigmoid curve -- I'd 
be interested in the physics behind this.

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(ep, aes(x = st_optmag, y = st_dist)) + geom_point(alpha = 0.2) + 
  coord_cartesian(ylim = c(0, 2500))

```

```{r echo=FALSE, warning= FALSE, message=FALSE}

ggplot(ep, aes(x = st_optmag, y = st_dist)) + geom_point(alpha = 0.2) + 
 coord_cartesian(ylim = c(0, 2500)) + scale_y_sqrt()

```


There appears to be some non-linear relationship between a star's distance from
earth and its optical magnitude. I'd guess this is a quadratic relationship 
because light propagates according to the inverse *square* law. Scaling the
y-axis by square root smooths out the relation a little bit, but
it's still not quite linear. 

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(ep, aes(x = factor(pl_pnum), y = pl_bmassj, group = factor(pl_pnum))) +
  geom_boxplot() + coord_cartesian(ylim = c(0, 20))


```

As the number of planets in a system increases, the mass of those planets
tends to decrease.



# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The relationship between planetary mass and planetary radius, as well as
between planetary mass and planetary density is not as positievly correlated as expected. One would expect a strong correlation in both scenarios, but there is
something going on in mass, radius, and density preventing this. It definitely
warrants further investigation. Adding additional variables might definitively
explain this.

There's a weak negative correlation between star distance from earth and 
planetary mass. I'd expect this to be explained by bias in different discovery methods. Perhaps the methods that are capable of discovering planets very far
from earth are biased towards discovering lower mass planets.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Adding another variable to the analysis is beginning to tell more of a story.
The scatter plot comparing planetary density and planetary mass shows what I 
believe are terrestrial planets hugging the x axis. I believe these planets are
terrestrial because they have low mass and high density, which is what one would
expect from a terrestrial planet. Terrestrial planets have low mass due to
physical restrictions on their radius.

Most stars in our dataset have about one solar mass. I'm interested to know
if there's something special about stars that have one solar mass to make 
planets more likely to orbit them, or if this is simply a result of the 
distribution of all stars' masses.

### What was the strongest relationship you found?

The relationship between star optical magnitude and the square root of
star distance is the strongest relation I found. I believe this is due to light
from a star propagating according to the inverse square of a star's distance. 

# Multivariate Plots Section


```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(ep, aes(x = st_dist, y = pl_bmassj, color = pl_discmethod)) +
  geom_point(alpha = 0.2) + scale_color_brewer(type='qual', palette = 2)

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(ep, aes(x = st_dist, y = pl_bmassj, color = pl_discmethod)) +
  geom_point(alpha = 0.2) + scale_color_brewer(type='qual', palette = 2) + 
  coord_cartesian(ylim = c(0, 25))

```


This plot has some fascinating qualities! Almost every single planet 
discovered more than 3000 light years away was discovered through microlensing!
Radial velocity seems to be the most common discovery method for relatively 
close planets (and seems to have no problem discovering high mass planets), 
while the transit method seems to easily discover planets a little further away,
but less than 2000 light years away. Planets discovered via transit also seem to
have lower mass than planets discovered via radial velocity. Planets discovered
via imaging are generally very near earth, and can have relatively high mass.

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(ep, aes(x = pl_radj, y = pl_bmassj)) +
  geom_point(size = .5*ep$pl_dens, alpha = 0.5) +
  coord_cartesian(xlim = c(0, 3), ylim =c(0, 10))

```

This plot gives us a better idea of the relationship between planetary radius,
mass, and density. Density is represented by the size of the points here,
making it apparent that planets with high density are clustered near 0 Jupiter
masses and 0 Jupiter radii. This cluster in the bottom left of the plot almost
certainly represents terrestrial planets.

We do see an interesting phenomenon around 1 solar mass, where many planets
with mass greater than 5 Jupiter masses have higher density. These planets
must have gaseous composition due to their radius and mass, but it's surprising
to see a gas planet with higher density than the numerous gas giants with much 
lower density below this phenomenon. What's going on with these planets?

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(ep, aes(x = pl_radj, y = pl_bmassj, color = factor(pl_pnum))) +
  geom_point(alpha = 0.2) + coord_cartesian(xlim = c(0, 3), ylim =c(0, 30))

```

Planets that are part of a system with only one planet tend to have higher 
radius than those that have multiple planets in their system. Could this mean
that terrestrial planets (which have lower radius than gaseous planets) are more
likely to be part of systems with multiple planets?

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(ep, aes(x = pl_orbsmax, y = pl_bmassj, color = pl_discmethod)) +
  geom_point(alpha = 0.2) + coord_cartesian(xlim = c(0, 7), ylim = c(0, 25))

```

The transit method doesn't discover many planets that are rotating more than 
0.5 AU from their host star. These planets tend to lie on a vertical line around 
0 AU. Let's investigate the properties of planets discovered by transit in the
next plot!

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(trans, aes(x = pl_dens, y = pl_bmassj, color = factor(pl_pnum))) +
         geom_point() + coord_cartesian(xlim = c(0, 15), ylim = c(0, 5))

```

Planets that have high mass, low density and that are part of systems with
only one planet are seen on the left side of this plot. Planets with higher 
density and lower mass hug the x-axis.


```{r echo=FALSE, warning=FALSE, message=FALSE}

m1 <- lm(ep, formula = pl_bmassj ~ pl_radj)
m2 <- lm(ep, formula = pl_bmassj ~ pl_radj + pl_dens)

m2 %>% summary()
```

This linear model accounts for only 27.2% of the variance of planetary mass!
We will explore this in the multivariate analysis section.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Radius, mass, and density generally behaved as predicted. As radius increases,
so does mass, while density decreases. As mass increases, density also increase,
and so does radius. 

Among planets discovered via the transit method, one group of planets has
low density, high mass, and only one planet in their system. The other (smaller)
group of planets has higher density, much lower mass, and more planets in their
system. I believe these two groups represent hot Jupiters and terrestrial
planets respectively.

Planets discovered via transit tend to be a moderate distance from earth (100
to 2000 light years away), have relatively low mass, and orbit close to their 
star.

Planets discovered via radial velocity tend to be close to earth, and have
very high mass. 

Planets discovered via microlensing tend to be very far from earth
(> 2000 light years), and have low mass.

Planets discovered via imaging tend to be very close to earth, and have low
mass.

### Were there any interesting or surprising interactions between features?

I'm surprised different discovery methods yield such distinct types of planets.
There is some overlap between radial velocity and imaging, but the other 
discovery methods have clear strengths and weaknesses in their ability to 
discovery planets. 

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I created a linear model that predicts mass of a planet based on its radius,
density, and mass of the star it orbits. This linear model accounted for only
27.6% of the variance in a planet's mass. This is somewhat counterintuitive 
because one would expect density and radius to mathematically define mass as
follows:

$$
\begin{equation}
volume = \frac{4}{3}\pi{r}^{3}\\
density = \frac{mass}{volume}
\end{equation}
$$

Which leads us to conclude

$$
\begin{equation}
mass = density*volume\\
mass = {density}*\frac{4}{3}\pi{r}^{3}
\end{equation}
$$

However, this is not a linear definition of mass (rather, it is cubic). Our 
*linear* model is not set up to predict mass in this way. If mass were
written as a linear combination of density and volume, our linear model
would be much better.

Perhaps more importantly, there are a few distinct groups of planets in our
dataset. Within these groups, I would expect a linear model to be much more
accurate than our current model. However, when we consider the dataset as a 
whole, we get a rather low correlation coefficient. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, warning=FALSE, message=FALSE, Plot_One}

hist_plot(ep, ep$pl_bmassj, .25) + coord_cartesian(xlim = c(0, 10)) + 
    labs(x = 'Jupiter masses', y = 'Number of planets', 
         title = 'Planetary Mass') + 
  theme(plot.title = element_text(hjust = 0.5))

```

```{r echo=FALSE, warning=FALSE, message=FALSE, Plot_one}

hist_plot(ep, ep$pl_bmasse, .5) + coord_cartesian(xlim = c(0, 5)) + 
    labs(x = 'Jupiter masses', y = 'Number of planets', 
         title = 'Planetary Mass') + 
  theme(plot.title = element_text(hjust = 0.5))

```

### Description One

The first plot shows us planets' distribution of mass on a larger scale (eg we 
can see very low mass planets and high mass planets in this plot). The second 


### Plot Two
```{r echo=FALSE, warning=FALSE, message=FALSE, Plot_Two}

ggplot(ep, aes(x = st_dist, y = pl_bmassj, color = pl_discmethod)) +
  geom_point(alpha = 0.2) + scale_color_brewer(type='qual', palette = 2) + 
  coord_cartesian(ylim = c(0, 25)) + labs(x = 'Star distance (light years)', 
  y = 'Jupiter masses', title = 'Planetary Mass by Star Distance', 
  color = 'Discovery method') + theme(plot.title = element_text(hjust = 0.5))

```

### Description Two
This plot allows us to see the strengths and limitations of the most common
discovery methods in regards to planetary mass and distance from earth.

* Radial velocity appears to be biased towards discovering planets near earth 
and with any mass from 0 to 20 Jupiter masses. 
* Transit most easily discovers planets between 100 and 1500 light years from
earth, with lower mass -- below 2.5 Jupiter masses. 
* Microlensing accounts for almost all planets discovered at a  distance greater
than 2000 light years away from earth. 
* Imaging appears to be biased towards planets near earth with high mass. 


### Plot Three
```{r echo=FALSE, warning=FALSE, message=FALSE, Plot_Three}

ggplot(trans, aes(x = pl_dens, y = pl_bmassj, color = factor(pl_pnum))) +
         geom_point() + coord_cartesian(xlim = c(0, 15), ylim = c(0, 5)) + 
  labs(x = 'Density (g/cm^3)', y= 'Jupiter masses', title = 'Density by
       Mass of Planets Discovered via Transit', color = 
      'Number of planets in system') +
  theme(plot.title = element_text(hjust = 0.5)) + 
  scale_color_brewer(type = 'div', palette = 1)

```

### Description Three

I believe we're seeing what astronomers refer to as hot Jupiters in this plot.
Hot Jupiters are gas giants that orbit very close to their stars. These are 
represented by the salmon colored data points with low density and high mass.
This leaves the terrestrial planets hugging the x-axis.

We can also see that planets with only one planet in the system tend to 
have higher mass than those with more than one planet in the system. This lends
support to the theory that terrestrial planets tend to have more planets in 
their systems, while gas giants have fewer.


------

# Reflection

The data on exoplanets was accumulated between 1992 and 2018, with most of the 
planets being discovered after 2008. There are 3736 observations with 142 
variables. I pared this down to 3696 observations with 17 variables after 
cleaning. 

The categorical variable of most interest to us is the discovery 
method of each planet. Each discovery method has different strengths and 
weaknesses -- the transit method shines in discovery of relatively close 
planets with moderate mass, while the radial velocity method specializes
in discovering very close and very massive planets.

The study and discovery of exoplanets is an extremely exciting field right now!
We're in the early stages of this field, and have very few data points compared
to how many planets likely exist in our galaxy, let alone the universe. 
Scientific methods may be able to approximate that average characteristics of
planets, but our current dataset is not enough to make large generalizations of
the distribution of any features of exoplanets.

For example, we could not say that there are more planets with high mass near
earth than far from earth, despite our data clearly showing that trend. Our 
methods for discovering planets near earth appear to be somewhat biased towards
discovering high mass planets, and our methods for discovering planets far from
earth tend not to be biased towards discovering high mass planets. 

This dataset suffers from lots of missing null values due to the data being
from several different sources (eg different exoplanet discovery missions). I 
left all the nulls in tact and went ahead with my analysis. This was a constant
limitation in this exploration.

One thing I'd love to do in the future with this dataset is to implement an 
algorithm that predicts what method a planet is likely to be discovered by
given a few of the planets hypothetical parameters. I believe this could be 
implemented by a multinomial logistic regression model.